Supplemental Materials for \\ Helixer: Cross-species Gene Annotation of Large Eukaryotic Genomes Using Deep Learning

### 2020-06-10 paper/main.tex

\abstract{
\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/Helixer} under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\  
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.
}

@@

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF 
(https://github.com/weberlab-hhu/GeenuFF). 
GeenuFF is a tool for checking and exploring genomic data, that 
stores all information inside a SQL database. Training and evaluation-ready data is 
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
is done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

@@

@@

\begin{table}[!t]
\processtable{Summary of Experimental Results\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 & 0.757 \\
Dilated CNN Baseline & 0.666 &  0.802 \\
Best Helixer model & 0.77  & 0.833  \\
Best model + varied input length & 0.834  & - \\
Best model (+ varied input length) + overlapping &  0.844  & 0.843  \\
(Varied input length +) overlapping + ensembling & 0.868  & 0.863  \\
\botrule
\end{tabular}}{{\it Note:} Values are the median in Subgenic F1 across all 
validation and test genomes of the respective group. Varied input length was only 
used in the animal case. The best model was chosen out of the eight models of the 
ensemble for having the best performance on the validation set of the training genomes.}
\end{table}

>>>>

\abstract{
\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/Helixer} under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\  
% normally contact is last author / person with permanent position (i.e. Andreas). 
% Not sure how important that is, but if we're skipping that, then we can change it back to Felix
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.
}

@@

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF 
\citep{denton2019}. GeenuFF is a tool for checking and exploring genomic data, that 
stores all information inside a SQL database. Training and evaluation-ready data is 
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
is done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

@@

During data generation, we query for the longest transcript of each gene and 
disregard FASTA sequences that have no functional annotation. GeenuFF checks the 
genomic annotations for potential errors during import and is able to mark those 
areas. We use this information to effectively blank out those bases during training 
by using the sample weights described in Table \ref{tab:arrays}. The vast majority 
of bases masked lie in the intergenic region as the most prevelant error is a 
missing UTR and GeenuFF marks a potentially large intergenic region for it. Table 
\ref{tab:statistics_all} shows statistics about the masking.

For the training itself, we divide each continous genomic sequence into 20.000 bases 
long subsequences, which are then used as input or label for the neural network 
together with the sample weights. We append zero padding if the subsequences are 
not long enough. If a subsequence is fully marked as erroneous, we do not include 
it anywhere.

@@

\begin{table}[!t]
\processtable{Summary of Experimental Results\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 & 0.757 \\
Dilated CNN Baseline & 0.666 &  0.802 \\
Best Helixer model & 0.77  & 0.833  \\
Best model + varied input length & 0.834  & - \\
Best model (+ varied input length) + overlapping &  0.844  & 0.843  \\
(Varied input length +) overlapping + ensembling & 0.868  & 0.863  \\
\botrule
\end{tabular}}{{\it Note:} Values are the median in Subgenic F1 across all 
validation and test genomes of the respective group. Varied input length was only 
used in the animal case. The best model was chosen out of the eight models of the 
ensemble for having the best performance in the reported setup.}
\end{table}

### 2020-06-10 paper/supplements.tex

\documentclass{article}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{amssymb}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
\newgeometry{vmargin={20mm}, hmargin={25mm}}

@@

\title{Supplemental Materials for \\ Helixer: Cross-species Gene Annotation of Large Eucaryotic Genomes Using Deep Learning}

@@

@@

%\renewcommand\thefigure{S7}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_002}}
\caption{
Four additional example helixer predictions in the context of RNAseq data, the reference,
and Augustus' prediction for {\it M. esculenta}. The examples were chosen so
that a) Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, b) Helixer had high accuracy but the
reference was not supported by the RNAseq data, c) Helixer had low accuracy
and the reference was supported by the RNAseq data, and d) helixer had low
accuracy but the reference was not supported by RNAseq data. Plotability was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii)
Augustus' predictions in matrix form, and iv) Helixer's predictions. The reference
and Augustus have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure}

\begin{figure}[!h]
\label{supfig:cov_example_03}
%\renewcommand\thefigure{S8}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_003}}
\caption{
Four additional example helixer predictions in the context of RNAseq data, the reference,
and Augustus' prediction for {\it M. esculenta}. The examples were chosen so
that a) Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, b) Helixer had high accuracy but the
reference was not supported by the RNAseq data, c) Helixer had low accuracy
and the reference was supported by the RNAseq data, and d) helixer had low
accuracy but the reference was not supported by RNAseq data. Plotability was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii)
Augustus' predictions in matrix form, and iv) Helixer's predictions. The reference
and Augustus have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure}

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{makecell}
\usepackage{amssymb}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
\newgeometry{vmargin={20mm}, hmargin={25mm}}

@@

\title{Supplemental Materials for \\ Helixer: Cross-species Gene Annotation of Large Eukaryotic Genomes Using Deep Learning}

@@

\newpage
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llll@{}}
Purpose&Tool&Version&Customized Parameters\\
Read quality control&FastQC&v0.11.5&\\ \\
Adapter, Quality trimming&Trimmomatic&0.36&\makecell[l]{ILLUMINACLIP:TruSeq3-PE-2.fa:3:30:10:1:true\\MAXINFO:36:0.7 MINLEN:36}\\ \\
Mapping&Hisat2&2.1.0&\makecell[l]{Both: --max-seeds 8 --dta\\Plants only: --pen-canintronlen G,-8,1.5 --pen-noncanintronlen G,-8,1.5} \\ \\
Bam conversion, sorting and indexing&Samtools&1.6&\\ \\
Mapping quality control&PicardTools&52.0&\makecell[l]{CollectRnaSeqMetrics\\STRAND=SECOND\_READ\_TRANSCRIPTION\_STRAND}\\ \\
Summarizing and organizing&MultiQC&1.8\\
\caption{List of bioinformatics software, versions, and parameters used in processing
public RNAseq data.}
\label{suptab:bioinfo_params}

@@

%\renewcommand\thefigure{S7}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_002}}
\caption{
Four additional example helixer predictions in the context of RNAseq data, the reference,
and Augustus' prediction for {\it M. esculenta}. The examples were chosen so
that a) Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, b) Helixer had high accuracy but the
reference was not supported by the RNAseq data, c) Helixer had low accuracy
and the reference was supported by the RNAseq data, and d) helixer had low
accuracy but the reference was not supported by RNAseq data. Feasibility for visualization was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii)
Augustus' predictions in matrix form, and iv) Helixer's predictions. The reference
and Augustus have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure}

\begin{figure}[!h]
\label{supfig:cov_example_03}
%\renewcommand\thefigure{S8}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_003}}
\caption{
Four additional example helixer predictions in the context of RNAseq data, the reference,
and Augustus' prediction for {\it M. esculenta}. The examples were chosen so
that a) Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, b) Helixer had high accuracy but the
reference was not supported by the RNAseq data, c) Helixer had low accuracy
and the reference was supported by the RNAseq data, and d) helixer had low
accuracy but the reference was not supported by RNAseq data. Feasibility for visualization was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii)
Augustus' predictions in matrix form, and iv) Helixer's predictions. The reference
and Augustus have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure}
